Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/01_Introduction.Rmd
restructuring the introduction

To answer these questions, a theoretical framework is required that permits the comparison of human behaviour across spatial, temporal and cultural boundaries: The assumption that the cremation aspect of Early Bronze Age burial rituals in Britain could be related to the same aspect in Hungary hundreds of years later is not self-evident. *Cultural evolution theory* provides such a framework [@creanzaCulturalEvolutionaryTheory2017]. The spread of ideas in human populations over space and time is one of its major topics: *Cultural transmission*.

@@

*Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution theory, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style]. Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their diffusion can be expected to increase over time if we assume reasonable human actors. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. They often serve archaeologists as a proxy for precise relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicSimulatingCulturalTransmission2014; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

@neimanStylisticVariationEvolutionary1995 takes a closer look at this processes and develops a simple framework for the simulation of non-functional idea propagation. He identifies the essential phenomena *drift* (random dominance of similar traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data.

The workflow suggested by Neiman is relevant to the burial rites data examined in this paper and was an important methodological inspiration: As burial traditions are ritualised behaviour which is most of the time heavily linked to religious and spiritual beliefs they can be assumed to be stylistic traits. Of course in reality, beyond the distinction of @dunnell1978style, most behaviour is not unambiguously stylistic or functional. Burial rites cause costs regarding working time and material to a varying extent. These expenses might have been influenced by or had their effect on the human-nature relation: Pyres, for example, require a significant amount of wood and the erection of barrows much workforce. Still, it can be expected that these functional disadvantages were not critical most of the time -- at least regarding the general decision which kind of burial tradition is applied. Most probably they influence the magnitude of invested energy for a burial, just like the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].

>>>>

Burial traditions are a ritual behaviour that is transmitted between generations. They are therefore suitable for study using methods of *Cultural transmission theory*, a special aspect of *Cultural evolution* concerned with the spread of ideas in human populations [@creanzaCulturalEvolutionaryTheory2017]. This theoretical framework permits the comparison of behaviour across spatial, temporal and cultural boundaries. *Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution theory, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style] which is very relevant for burial traditions: Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their diffusion can be expected to increase over time if we assume reasonable human actors. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. They serve archaeologists as a proxy for precise relative chronology. Burial traditions can be descriped as a special form of *stylistic traits*. Special because of their strong dependence on religious and spiritual beliefs and their connection with the emotionally stirring topic of death, but still subject to seemingly random fashion changes.

Of course in reality most behaviour is not unambiguously stylistic or functional. Burial rites cause costs regarding working time and material to a varying extent. These expenses might have been influenced by or had their effect on the human-nature relation: Pyres, for example, require a significant amount of wood and the erection of barrows much workforce. Still, it can be expected that these functional disadvantages were not critical most of the time -- at least regarding the general decision which kind of burial tradition is applied. Most probably they influence the magnitude of invested energy for a burial, just like the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].

Dunnell postulates that the distribution and emergence of stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicSimulatingCulturalTransmission2014; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001]. @neimanStylisticVariationEvolutionary1995 takes a closer look at these processes and develops a simple framework for the simulation of non-functional idea propagation. He identifies the essential phenomena *drift* (random dominance of similar traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data. This workflow was an important methodological inspiration for the burial rites data examination in this paper.

@@
